Omics-based Single Step Trait Prediction

### 2017-09-17 paper/manuscript.tex
paper, first full abstract draft

\begin{abstract}
Predicting the performance of candidates is indispensable for selection
decisions in animal and plant breeding.
Novel `omics' predictors have shown to improve upon genomics-based predictions
of some complex traits but are yet expensive to sample.
Here we show that single step prediction can be applied to a quantitative
predictor.

Using two maize datasets of inbred lines and hybrids we show that transcriptomic
data can successfully be imputed, boosting forecasts when their predictive
ability exceeds that of pedigree or genomic data.

Equipping only a subset of genotypes with `omics' predictors and imputing all
others using pedigree or genomic data could enable breeders to improve trait
predictions while keeping costs manageable.
Employing `omics' predictors could particularly benefit candidate selection in
hybrid breeding.
\end{abstract}

% Please add here a significance statement to explain the relevance of your work
\section{Article Summary}
$\dots$

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either
  Exp1 or Exp2.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar shows that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed (light gray fraction).
  The four bars to the right outline the principle of generating core sets for
  Exp1.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the white symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor whereas the remaining $100\% - x\%$ of genotypes
  are declared to be covered only by the `complete' predictor.
  }
\label{fig:compound-plot}
\end{figure}

@@

\Genetics2level{Exp2}
The maize hybrid data set encompasses seven agronomic traits and three
predictors, namely pedigree data (P), genomic information (G), transcriptomic
information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Reduced'.
In the `Reduced' set, transcriptomic data were the best single predictor for
ADL, PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the reduced set, P had the highest predictive ability for STA and DMC\@.
For FAT and SUG, G was the superior predictor inside the reduced set.
In the full set of 1,521 hybrids, predictive abilities obtained with G increased
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information
($\tensor[_{C}]{G}{}\tensor[_{I}]{T}{}$) in a single-step model yielded a small
improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the reduced set.
Predictive abilities resulting from combinations of pedigree with genomic
($\tensor[_{C}]{P}{}\tensor[_{I}]{G}{}$) information were better than P alone
for all traits and $\tensor[_{C}]{P}{}\tensor[_{I}]{T}{}$ improved upon P in
the `full' set for four traits.

A previous study using the same data as in Exp2 concluded that modeling of
specific combining ability (SCA) effects did not appreciably improve predictive
abilities over that of a purely additive model for any of the investigated
traits.
Nevertheless, the inclusion of SCA effects in a single step prediction model
would be as straightforward as computing the Kronecker product between extension of single step prediction

>>>>

\begin{abstract}
Predicting genotypic values with great precision is pivotal for effective
candidate selections in animal and plant breeding.
Novel `omics' predictors have shown to improve upon established genome-based
predictions of some complex traits but are yet expensive to sample.
In animal breeding, single step prediction was introduced to efficiently use
pedigree information, collected on a large number of animals, and genomic
information, collected on a smaller subset of animals, simultaneously without
bias.
Single step prediction has not been attempted with a predictor whose features
display heritabilities smaller than one, yet.
Using two maize datasets of inbred lines and hybrids we show that transcriptomic
data can be imputed successfully, boosting forecasts when their predictive
ability exceeds that of pedigree or genomic data.
Covering only a subset of genotypes with `omics' predictors and imputing all
others using pedigree or genomic data could enable breeders to improve trait
predictions while keeping costs manageable.
Employing `omics' predictors could particularly benefit candidate selection in
hybrid breeding where the success of forecasts is a strongly convex function of
predictive ability.
\end{abstract}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either
  Exp1 or Exp2.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar shows that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed (light gray fraction).
  The four bars to the right outline the principle of generating core sets for
  Exp1.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the white symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor (dark gray) whereas the remaining $100\% - x\%$ of
  genotypes are declared to be covered only by the `complete' predictor (light
  gray).
  }
\label{fig:compound-plot}
\end{figure}

@@

\Genetics2level{Exp2}
The maize hybrid data set encompasses seven agronomic traits and three
predictors, namely pedigree data (P), genomic information (G), transcriptomic
information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Reduced'.
In the `Reduced' set, transcriptomic data were the best single predictor for
ADL, PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the reduced set, P had the highest predictive ability for STA and DMC\@.
For FAT and SUG, G was the superior predictor inside the reduced set.
In the full set of 1,521 hybrids, predictive abilities obtained with G increased
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information
($\tensor[_{C}]{G}{}\tensor[_{I}]{T}{}$) in a single-step model yielded a small
improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the reduced set.
Predictive abilities resulting from combinations of pedigree with genomic
($\tensor[_{C}]{P}{}\tensor[_{I}]{G}{}$) information were better than P alone
for all traits and $\tensor[_{C}]{P}{}\tensor[_{I}]{T}{}$ improved upon P in
the `full' set for four traits.
A previous study using the same data as in Exp2 concluded that modeling of
specific combining ability (SCA) effects did not appreciably improve predictive
abilities over that of a purely additive model for any of the investigated
traits.
Nevertheless, the inclusion of SCA effects in a single step prediction model
would be as straightforward as computing the Kronecker product between the
feature matrices $\mathbf{M}_{D}$ and $\mathbf{M}_{F}$ of the two corresponding
heterotic groups~\cite{Martini2016}.
